---
title: "How do we set up a Python virtual environment in the R shinyapps.io server?"
date: "2023-02-09"
categories:
  - Technichal
editor: visual
---

Today, I will introduce what are the initial steps that you need to follow when setting up a Python virtual environment in R for the shinyapps.io server.

**Note: You need to have an account in shinyapps.io server. The page can be find in [here](https://www.shinyapps.io/).**

First, you need to create `.Rprofile` file within the directory of your shiny app. Then, initialize a suitable name for the $\color{red}{\text{Python virtual environment}}$.

```{r}
#| warning: false
#| message: false
#| eval : false
VIRTUALENV_NAME = "new_env"
```

After that, we need to set environment variables in `.Rprofile` file as follows.

```{r}
#| warning: false
#| message: false
#| eval : false
Sys.setenv(PYTHON_PATH = 'python3')
# Installs into default shiny virtualenvs dir
Sys.setenv(VIRTUALENV_NAME = VIRTUALENV_NAME) 
Sys.setenv(RETICULATE_PYTHON = paste0('/home/shiny/.virtualenvs/', 
                                      VIRTUALENV_NAME, '/bin/python'))
```

The next step is to create the $\color{red}{\text{Python virtual environment}}$. To do that, you can write the following code chunks in `server.R` or your shiny app script.

In there, first of all, you need to get the environment variables.

```{r}
#| warning: false
#| message: false
#| eval : false
virtualenv_dir = Sys.getenv("VIRTUALENV_NAME")
python_path = Sys.getenv("PYTHON_PATH")
```

Next, create a Python virtual environment by specifying the Python path.

```{r}
#| warning: false
#| message: false
#| eval : false
reticulate::virtualenv_create(virtualenv_dir, python = python_path)
```

Then, you have to install Python dependencies. To do this, you can install the packages by directly specifying them.

```{r}
#| warning: false
#| message: false
#| eval : false
reticulate::virtualenv_install(virtualenv_dir, packages = c("pandas==1.3.5")) 
```

If not, can use **requirement.txt** which contains all the packages.

```{r}
#| warning: false
#| message: false
#| eval : false
reticulate::virtualenv_install(virtualenv_dir, c("-r", "requirements.txt"))
```

Finally, define the Python virtual environment to be used used by `reticulate`.

```{r}
#| warning: false
#| message: false
#| eval : false
reticulate::use_virtualenv(virtualenv_dir, required = TRUE)
```
