---
title: "Can the tSNE results be reproducible?"
date: "2023-02-05"
categories:
  - tSNE
editor: visual
---

If you are an R user struggling to access **HARPS GTO sample dataset** and have the question that **Can't we reproduce the t-distributed stochastic neighbor embedding (tSNE) results that already exist?** In their mind, this blog is for you. Today, I will introduce my workflow of accessing processed data from original data, downloading the processed data, and data processing to initialize cluster labels and show my surprising findings on reproducing tSNE results.

This example is relevant to the paper, [**"Dissecting stellar chemical abundance space with t-SNE" (Anders et al. 2018, Astronomy & Astrophysics 619, A125)**](https://arxiv.org/abs/1803.09341).

For simplicity, I added all the codes separately and linked them through the process.

Please note that you will need to $\color{green}{\text{install python library}}$ called `open_data`, which you can find in [open_data.py](https://gist.github.com/JayaniLakshika/602b4867582fd047293700ab720ef472) as the initial step.

## Data processing steps already followed by open_data library

The original data can be found in the GitHub repository of the authors as [DelgadoMena2017.fits](https://github.com/fjaellet/abundance-tsne/blob/master/data/DelgadoMena2017.fits), which includes some existing cluster labels and abundance determination for $\color{blue}{\text{Mg, Al, Si, Ca, TiI, Fe, Cu, Zn, Sr, Y, ZrII, Ce, and Ba}}$. This file contains details regarding $\color{red}{\text{1059}}$ stars.

Then, how do we get only $\color{red}{\text{530}}$ stars? ðŸ¤”

Let's investigate...

According to the authors, the sample needed to be analyzed in a more restricted temperature range to obtain reliable tSNE abundance maps. The reason is that specific abundance trends dominate underlying temperature trends. Therefore,

1.  choose an effective temperature range of $\color{brown}{5300 \text{ K} < T_{eff} < 6000 \text{ K}}$.

Only $\color{green}{539}$ stars were satisfied in this step. Next,

2.  exclude stars with $\color{brown}{\text{log } g_{HIP}} < 3$ which remove $\color{green}{\text{one}}$ star.

Then,

3.  select successful abundance determination for $\color{brown}{\text{Mg, Al, Si, Ca, TiI, Fe, Cu, Zn, Sr, Y, ZrII, Ce, and Ba}}$ which use as input for tSNE.

Only $\color{green}{533}$ stars have remained because others contain missing values. Furthermore,

4.  to compensate for the fact that tSNE does not take into account individual (heteroscedastic) uncertainties in the data, the authors followed the approach of [Hogg et al. (2016)](https://iopscience.iop.org/article/10.3847/1538-4357/833/2/262/meta) and rescaled each abundance by the median uncertainty in that element, assuming an abundance uncertainty floor of 0.03 dex. Additionally,

5.  remove stars that did not converge by using the age determination code, `StarHose code`.

The $\color{green}{3}$ stars are discarded.

There are only $\color{green}{530}$ stars in our final sample. ðŸ‘

***Note: You do not need to consider the above steps because the `open_data` package has already done it for us.***

## Our workflow

Let's begin... ðŸ¤“

1.  To download the processed data as a CSV file and save it locally, you need to run [download_original.py](https://gist.github.com/JayaniLakshika/4d646a2ad364b1a97ff8aaf050a6d568),

2.  Run [this code](https://gist.github.com/JayaniLakshika/3215e7b7c72e2d2c3424d477ba2fb296) to process data further as indicated in the original paper to compute missing variables, new variables, cluster labels and produce plots of authors t-SNE results,

3.  Run [this code](https://gist.github.com/JayaniLakshika/b1e440a8432fd33c3587cf16e04cf8f7) to conduct our own t-SNE results.

Here, I used the $\color{red}{\text{Perplexity parameter as 40}}$ authors already mentioned that in the paper.

Here is the output..ðŸ¤—

```{r}
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(readr)
library(ggplot2)
library(plotly)
library(patchwork)
library(Rtsne)

data <- read_csv("/Users/jpiy0001/Desktop/PhD Monash research files/Meetings/Meeting_24/HARPS_GTO/Process_data/HARPS_data_by_author.csv")

data <- data %>% mutate(vXg_minus = vXg * (-1),
                YMg = YFe - MgFe,
                AlMg = AlFe - MgFe,
                YAl = YFe - AlFe,
                SrBa = SrFe - BaFe,
                ZrBa = ZrIIFe - BaFe,
                CFe = `[C/H]_SA17` - feh,
                CO = `[C/H]_SA17` - `[O/H]_6158_BdL15`,
                sqrt_Vxg_Vzg = sqrt(vXg**2. + vZg**2. ),
                MgY = MgFe - YFe,
                BaY = BaFe - YFe,
                Rg_Rc = Rg - Rc,
                Rc_rb = Rc - rb,
                sqrt_errY_errMg = sqrt(errY**2. + errMg**2.),
                sqrt_errAl_errMg = sqrt(errAl**2. + errMg**2.),
                sqrt_errY_errAl = sqrt(errY**2. + errAl**2.),
                sqrt_errSr_errBa = sqrt(errSr**2. + errBa**2.),
                sqrt_errZrII_errBa = sqrt(errZrII**2. + errBa**2.),
                sqrt_errCH_errfeh = sqrt(`e_[C/H]_SA17`**2. + erfeh**2.),
                sqrt_errCH_errOH = sqrt(`e_[C/H]_SA17`**2. + `e_[O/H]_6158_BdL15`**2.),
                sqrt_vXg_sig_vZg_sig = sqrt(vXg_sig**2. + vZg_sig**2.),
                sqrt_errY_errBa = sqrt(errY**2. + errBa**2.),
                sqrt_Rg_sig_Rc_sig = sqrt(Rg_sig**2. + Rc_sig**2.),
                sqrt_rb_err_Rc_sig = sqrt(rb_err**2. + Rc_sig**2.),
                abs_JR_st_U_JR_st_L = 0.5*abs(JR_st_U-JR_st_L),
                abs_JZ_st_U_JZ_st_L = 0.5*abs(JZ_st_U-JZ_st_L))

## Assign outer cluster names
data <- data %>%
mutate(outer_cluster_teffcut40 = ifelse(tsne_class_teffcut40 == "thin", "",
                             ifelse(tsne_class_teffcut40 == "thick1", "",
                             ifelse(tsne_class_teffcut40 == "thick2", "",
                             ifelse(tsne_class_teffcut40 == "thick3", "",
                             ifelse(tsne_class_teffcut40 == "thick4", "",
                             ifelse(tsne_class_teffcut40 == "mpthin", "",
                             ifelse(tsne_class_teffcut40 == "mpthintrans", "Transition group",
                             ifelse(tsne_class_teffcut40 == "smr", "",
                             ifelse(tsne_class_teffcut40 == "t4trans", "",
                             ifelse(tsne_class_teffcut40 == "youngthin", "Young local disc",
                             ifelse(tsne_class_teffcut40 == "debris1", "",
                             ifelse(tsne_class_teffcut40 == "debris2", "",
                             ifelse(tsne_class_teffcut40 == "debris3", "[s/Fe]-enhanced",
                             ifelse(tsne_class_teffcut40 == "debris4", "",
                             ifelse(tsne_class_teffcut40 == "debris5", "",
                             ifelse(tsne_class_teffcut40 == "smr2", "",
                             ifelse(tsne_class_teffcut40 == "t2trans1", "Debris candidate",
                             ifelse(tsne_class_teffcut40 == "highTi", "Extreme-Ti star",
                             ifelse(tsne_class_teffcut40 == "lowMg", "Low-[Mg/Fe] star",
                             ifelse(tsne_class_teffcut40 == "highAlMg?", "High-[Al/Mg] star",

                                     "")))))))))))))))))))))

data <- data %>%
  mutate(outer_cluster_teffcut40_mc = ifelse(tsne_class_mc == "thin", "Thin Disc",
                                   ifelse(tsne_class_mc == "thick1", "Thick Disc I",
                                                 ifelse(tsne_class_mc == "thick2", "Thick Disc II",
                                                        ifelse(tsne_class_mc == "thick3", "Inner Disc I",
                                                               ifelse(tsne_class_mc == "thick4", "Inner Disc II",
                                                                      ifelse(tsne_class_mc == "mpthin", "Outer Thin Disc",
                                                                             ifelse(tsne_class_mc == "smr", "SMR",
                                                                                    ifelse(tsne_class_mc == "t4trans", "Inner Disc III",
                                                                                           ifelse(tsne_class_mc == "debris1", "",
                                                                                                  ifelse(tsne_class_mc == "debris2", "",
                                                                                                         ifelse(tsne_class_mc == "debris3", "Satellite debris",
                                                                                                                ifelse(tsne_class_mc == "debris4", "",
                                                                                                                       ifelse(tsne_class_mc == "debris5?", "",
                                                                                                                              ifelse(tsne_class_mc == "t2trans1", "TII/III",
                                                                                                                                     ifelse(tsne_class_mc == "t2trans2", "",
                                                                                                                                            ifelse(tsne_class_mc == "highTi", "Extreme-Ti star",
                                                                                                                                                   ifelse(tsne_class_mc == "thicklow", "Lowest-[Fe/H] star",


                                                                                                                                                                               ""))))))))))))))))))


data <- data %>%
  mutate(outer_cluster_teffcut40_errlim_mc = ifelse(tsne_class_errlim_mc == "thin", "Thin Disc",
                                                    ifelse(tsne_class_errlim_mc == "thick1", "Thick Disc",
                                                           ifelse(tsne_class_teffcut40 == "thick2", "",
                                                                  ifelse(tsne_class_teffcut40 == "thick3", "",
                                                                         ifelse(tsne_class_teffcut40 == "thick4", "Metal-rich Thick Disc",
                                                                                ifelse(tsne_class_teffcut40 == "mpthin", "Metal-poor Thin Disc",
                                                                                       ifelse(tsne_class_teffcut40 == "youngthin", "Young Local Disc",
                                                                                              ifelse(tsne_class_teffcut40 == "smr", "SMR",
                                                                                                     ifelse(tsne_class_teffcut40 == "t4trans", "Transition",

                                                                                                                   ifelse(tsne_class_teffcut40 == "debris1", "Debris",
                                                                                                                          ifelse(tsne_class_teffcut40 == "debris2", "",
                                                                                                                                 ifelse(tsne_class_teffcut40 == "debris3", "",
                                                                                                                                        ifelse(tsne_class_teffcut40 == "debris4", "",
                                                                                                                                               ifelse(tsne_class_teffcut40 == "debris5?", "",
                                                                                                                                                      ifelse(tsne_class_teffcut40 == "t2trans1", "",
                                                                                                                                                             ifelse(tsne_class_teffcut40 == "t2trans2", "",
                                                                                                                                                                    ifelse(tsne_class_teffcut40 == "highTi", "",
                                                                                                                                                                           ifelse(tsne_class_teffcut40 == "t2trans3", "",
                                                                                                                                                                                  ifelse(tsne_class_teffcut40 == "smr2", "",
                                                                                                                                                                                         ifelse(tsne_class_teffcut40 == "lowMg", "",

                                                                                                                                                                                         "")))))))))))))))))))))





data <- data %>%
  mutate(outer_cluster_teffcut40_simple = ifelse(tsne_class_teffcut40_simple == "thin", "Thin Disc",
                                          ifelse(tsne_class_teffcut40_simple == "thick1", "Thick Disc I",
                                                 ifelse(tsne_class_teffcut40_simple == "thick2", "Thick Disc II",
                                                        ifelse(tsne_class_teffcut40_simple == "thick3", "Thick Disc III",

                                                                      ifelse(tsne_class_teffcut40_simple == "mpthin", "Metal-poor thin disc",

                                                                                    ifelse(tsne_class_teffcut40_simple == "smr", "SMR",
                                                                                           ifelse(tsne_class_teffcut40_simple == "t1trans", "Transition I",

                                                                                                         ifelse(tsne_class_teffcut40_simple == "debris", "Satellite debris",
                                                                                                                ifelse(tsne_class_teffcut40_simple == "highAlMg", "High-[Al/Mg]",
                                                                                                                       ifelse(tsne_class_teffcut40_simple == "t3trans", "Transition III",
                                                                                                                              ifelse(tsne_class_teffcut40 == "highTioutlier", "Extreme-Ti star",
                                                                                                                                     ifelse(tsne_class_teffcut40_simple == "lowalphaoutlier", "Low-[Mg/Fe] star",


                                                                                                                                                                               "")))))))))))))







### Actual result generation

actual_tsne <- data %>% select("Star","X_tsne_teffcut40", "Y_tsne_teffcut40", "tsne_class_teffcut40", "outer_cluster_teffcut40",
                "tsne_class_teffcut40_simple", "outer_cluster_teffcut40_simple","X_tsne_teffcut40_errlim_mc", "Y_tsne_teffcut40_errlim_mc",
                "tsne_class_errlim_mc","outer_cluster_teffcut40_errlim_mc", "X_tsne_teffcut40_nofeh_mc", "Y_tsne_teffcut40_nofeh_mc",
                "X_tsne_teffcut40_mc", "Y_tsne_teffcut40_mc", "tsne_class_mc", "outer_cluster_teffcut40_mc")

actual_tsne1 <- actual_tsne %>% filter(outer_cluster_teffcut40 != "")#

actual_tsne1$cluster <- actual_tsne1$outer_cluster_teffcut40

actual_tsne2 <- actual_tsne %>% filter(outer_cluster_teffcut40 == "")

actual_tsne3 <- actual_tsne2 %>%  filter(outer_cluster_teffcut40_mc != "")

actual_tsne3 <- actual_tsne3 %>%
  mutate(outer_cluster_teffcut40_mc = ifelse(outer_cluster_teffcut40_mc == "SMR", "Thin Disc",
                                                    ifelse(outer_cluster_teffcut40_mc == "Lowest-[Fe/H] star", "Thick Disc I", outer_cluster_teffcut40_mc))) #

actual_tsne3$cluster <- actual_tsne3$outer_cluster_teffcut40_mc


stars <- c(actual_tsne1$Star, actual_tsne3$Star)

actual_tsne4 <- actual_tsne[-which(actual_tsne$Star %in% stars),]

actual_tsne5 <- actual_tsne4 %>% filter(outer_cluster_teffcut40_simple != "") #

actual_tsne5$cluster <- actual_tsne5$outer_cluster_teffcut40_simple

actual_tsne6 <- actual_tsne4 %>% filter(outer_cluster_teffcut40_simple == "")

actual_tsne6$cluster <- rep("Debris candidate", nrow(actual_tsne6)) #

data_with_cluster_label <- bind_rows(actual_tsne1, actual_tsne3, actual_tsne5, actual_tsne6)


tSNE_plot_actual <- data_with_cluster_label %>%
  ggplot(aes(x = X_tsne_teffcut40, y = Y_tsne_teffcut40, color = cluster)) +
  geom_point(aes(shape=cluster, color=cluster)) +
  scale_shape_manual(values=c(16, 5, 5, 13, 8, 15, 16, 15, 10, 17, 24, 25, 17, 17)) + ggtitle("Visualization from author's results by tSNE") + xlab("t-SNE X dimension") + ylab("t-SNE Y dimension") + coord_fixed()
```

```{r}
#| echo: false
#| warning: false
#| message: false

data1 <- read_csv("/Users/jpiy0001/Desktop/PhD Monash research files/Meetings/Meeting_24/HARPS_GTO/Process_data/HARPS_GTO_processed_with_cluster_label.csv")


selected_columns1 <- colnames(data1)[grepl("Mg|Al|Si|Ca|TiI|fe|Cu|Zn|Sr|Y|ZrII|Ce|Ba",colnames(data1))]

selected_columns_new1 <- selected_columns1[! selected_columns1 %in% c('Y_tsne_teffcut40', 'Y_tsne_teffcut40_old', 'Y_tsne_teffcut40_errlim_mc', "Y_tsne_teffcut40_errlim_mc_old", "Y_tsne_teffcut40_nofeh_mc", "Y_tsne_teffcut40_mc", "Yg_sig", "vYg", "vYg_sig", "[Fe/H]_C11", "Yg", "nCu", "nZn" , "nSr", "nY", "nZrII", "nBa", "nCe", "nAl", "nMg", "nSi", "nCa", "[Fe/H]_T13")]

selected_columns_new1 <- selected_columns_new1 %>% append("cluster") #"tsne_class_teffcut40",



#
selected_columns_new1 <- selected_columns_new1[selected_columns_new1 %in% c("feh", "AlMg", "TiIFe", "SiFe", "MgFe", "ZnFe", "CuFe", "YMg", "BaFe", "CeFe", "ZrBa", "SrBa", "cluster", "ZrIIFe", "CaFe", "AlFe", "YFe", "ZrFe", "SrFe")] ###Some what correct
#




df <- data1 %>% select(all_of(selected_columns_new1))



set.seed(20210091)
tSNE_fit <- df %>%
  select(where(is.numeric)) %>%
  #Rtsne(perplexity = 40, pca = FALSE, pca_center = FALSE, normalize = FALSE)
Rtsne(perplexity = 40)

#df_na_omit <- df %>% na.omit()

tSNE_df <- tSNE_fit$Y %>%
  as.data.frame()  %>%
  mutate(ID=row_number())

names(tSNE_df)[1:2] <- c("tSNE1", "tSNE2")

tSNE_df_plot <- tSNE_df %>%
  ggplot(aes(x = tSNE2,
             y = tSNE1,
             color = df$cluster)) +
  geom_point(aes(shape=df$cluster, color=df$cluster)) +
  scale_shape_manual(values=c(16, 5, 5, 13, 8, 15, 16, 15, 10, 17, 24, 25, 17, 17)) +  ggtitle("Visualization from generated results by tSNE") + xlab("t-SNE X dimension") + ylab("t-SNE Y dimension") +
  theme(legend.position = "none") + coord_fixed()


lemon::grid_arrange_shared_legend(tSNE_plot_actual, tSNE_df_plot, ncol = 2, position='bottom')
```

## Acknowledge

I am grateful for Prof. Dianne Cook who suggested me as an exercise for me.
